---
title: "Exploratory Time Series Analysis "
author: "Inês Silva, Maria Miguel Ribeiro, Renatha Vieira"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R and Packages
```{r include=FALSE}
library(astsa)
library(fpp2)
library(forecast)
library(ggseas)
library(ggplot2)
library(tsibble)
```


### Read the data into R with the following script

```{r}
monthly_data <- read.csv(file.choose())
head(monthly_data)
```

### Convert the data to time series

```{r}
mytimeseries <- ts(monthly_data[,-1], start=2006, end=2014, frequency=12)

```

### Construct time series plot

```{r}
autoplot(mytimeseries, facets=TRUE)
```



# Time plots and time series components


We start by understanding the main components of a time series and how to best represent the data. 
  
```{r}
str(mytimeseries)
tsplot(mytimeseries, ylab="NUM_OCCURR", main="NYC crime rate from 2004 to 2014")
```


 

#Month plots of the time series

```{r}
ggmonthplot(mytimeseries)+
  ylab("NUM_OCURR") +
  ggtitle("Seasonal plot: number of crime occurencies per month ")
```

#Log Transformation of the data
```{r}
par(mfrow=c(2,1))
autoplot((mytimeseries)) +
  ggtitle("NYC crime occurrencies from 2006 to 2014") +
  ylab("NUM_OCURR") +
  xlab("CMPLNT_FR_DT")
autoplot(log(mytimeseries)) +
  ggtitle("Log NYC crime occurrencies from 2006 to 2014") +
  ylab("log ($ CMPLNT_FR_DT)") +
  xlab("NUM_OCURR")
```


#Performing Box-Cox  transformation in the data

```{r}
mytimeseries.lambda=BoxCox.lambda(mytimeseries)
mytimeseries.lambda
mytimeseries.BC=BoxCox(mytimeseries,lambda=mytimeseries.lambda)
autoplot(mytimeseries.BC) +
  ggtitle("Box-Cox NYC crime occurrencies from 2006 to 2014") +
  ylab("NUM_OCURR") +
  xlab("CMPLNT_FR_DT")
```

#Diferencing data and keep it in a variable for further use
```{r}
my_log_data <- log(mytimeseries)
```

#Lagplots of the time series
```{r}
lag1.plot(mytimeseries,12)
lag1.plot(my_log_data,12)

```
# Notice that between lag 1 and lag 5 the correlation increases, than decreases until lag 12! This indicates a seasonal pattern in the data

#Using a linear model to fit the trend
```{r}
summary(fit <- lm(my_log_data~time(my_log_data))) # regress NUM_OCURR on time
```

#Detrended data using a linear model aproach
```{r}
par(mfrow=c(2,1))
tsplot(my_log_data, ylab="NUM_OCURR", col=3, lwd=2)
abline(fit)           # add the fitted regression line to the plot
tsplot(resid(fit), main="detrended")
lag1.plot(resid(fit),12)
```
#Using a difference operator aproach to remove the trend

Another approach is to filter the time series with the **difference operator**, $\nabla= 1-B$ where $B$ is the **lag operator** $B x_t=x_{t-1}$. So $$\nabla x_t = (1-B ) x_t =x_t -B x_{t} = x_t- x_{t-1}$$

The resulting time series $y_t=\nabla x_t$ represents the increments or change of $x$ on consecutive time points. Take the number of occurencies, $x_t$. Then $y_t = x_t- x_{t-1}$ represents the monthly increase of price: from January to February, February to March, etc.

```{r}
par(mfrow=c(2,1))
tsplot(my_log_data, ylab="NUM_OCURR", col=4, lwd=2)
tsplot(diff(my_log_data), ylab="NUM_OCURR", col=4, lwd=2, main="Number crime of occurencies differenced")
```

#Representing the respective lagplots of detrended time series
```{r}
lag1.plot(diff((my_log_data)),12)
```


#Smoothing methods: estimating the trend

```{r}
w <- c(0.1, rep(1, 6), 0.5) / 6
timeseries <- filter(my_log_data, sides = 2, filter = w)

# Plot the original data with gray color
plot(my_log_data, col = rgb(0.5, 0.6, 0.85, 0.9), ylim = c(0, 5), main = "Time Series Plot")

# Add lines with different colors and adjusted line widths
lines(timeseries, lwd = 2, col = "blue")  # Blue line for MA
lines(ksmooth(time(my_log_data), my_log_data, "normal"), lwd = 2, col = "green")  # Green line for kernel smooth
lines(lowess(my_log_data), lty = 1, lwd = 2, col = "orange")  # Orange line for Lowess

# Add legend with corresponding colors and adjusted line widths
legend(x = 2006, y = 50, legend = c("MA", "Kernel", "Lowess"), 
       col = c("blue", "green", "orange"), lwd = 2, text.col = c(4, 3, 6))
```



#Lag plots of the time series using forecast
```{r}
gglagplot(my_log_data)
gglagplot(diff(my_log_data))
```

#Removing trend and seasonality
```{r}
fit.times <- tslm(my_log_data ~ trend + season)
summary(fit.times)
```

```{r}
autoplot(my_log_data, series="Data") +
  autolayer(fitted(fit.times), series="Fitted") +
  xlab("CMPLNT_FR_DT") + ylab("NUM_OCURR") +
  ggtitle("NYC  crime rate")
tsplot(residuals(fit.times), main = "Residuals of NYC crime data")
```
 


#Using seasonal difference operator to remove the seasonality

```{r}
par(mfrow=c(2,1))
tsplot(my_log_data)
tsplot(diff(my_log_data,12))
monthplot(diff(my_log_data,12))
```
```{r}
diff_data_trend <- diff(my_log_data)
diff_data_seasonal <- diff(my_log_data, 12)
```

#Representing the respective lag plots
```{r}
lag1.plot(diff_data_trend,12)
lag1.plot(diff_data_seasonal,12)
```
#data com trend removida apresenta correlaçao negativa entre o lag 1 e o lag 12, ou seja entre um ano e outro
 



#ACF e PACF of ts detrended and with ts deseasonalized
````{r}
acf(diff_data_trend, main = "Autocorrelation Function (ACF)of Detrended Time Series")

acf(diff_data_seasonal, main = "Autocorrelation Function (ACF) of Deseasonalized Time Series")


pacf(diff_data_trend, main = "Partial Autocorrelation Function (PACF) of Detrended Time Series ")
pacf(diff_data_seasonal, main = "Partial Autocorrelation Function (PACF) of Deseasonalized Time Series")


```
#STL decompostion
```{r}
my_log_data.stlper=stl(my_log_data, s.window="periodic")
my_log_data.stl=stl(my_log_data, s.window=12)
plot(my_log_data.stlper)
plot(my_log_data.stl)
```
```{r}
trend_component <- my_log_data.stl$time.series[, "trend"]
seasonal_component <- my_log_data.stl$time.series[, "seasonal"]
residual_component <- my_log_data.stl$time.series[, "remainder"]
```

#Trend and Seasonal components
```{r}
lag1.plot(trend_component,12)
lag1.plot(seasonal_component,12)
```

#The remainder behaviour

```{r}
lag1.plot(residual_component,12)
autoplot(residual_component)
```
```{r}
acf(residual_component, main = "Autocorrelation Function (ACF)of Residuals")

pacf(residual_component, main = "Partial Autocorrelation Function (PACF) Residuals ")

```



#Augmented Dickie Fuller Test (ADF)

```{r}
#install.packages("tseries")
library(tseries)
result <- adf.test(my_log_data)
print(result)
```

#Train test split using the original dataset withou no transformations or differentiations. The original data is saved in the variable 'my_log_data' which refers to the data with log transformation only

```{r}
library(forecast)


y_train <- window(my_log_data, start = 2006, end=2012+6/12)
y_test <- window(my_log_data, start= 2012+6/12, end = 2014)
#Modelo ARIMA sazonal com dados de treino
model_arima_sazonal <- auto.arima(y_train)
summary(model_arima_sazonal)
```
```{r}
model2 <- arima(y_train, order=c(2,1,2), seasonal=list(order=c(1,1,2), period=12))

# Visualize o resumo do modelo
summary(model2)

```

```{r}

acf(y_train, main = "Autocorrelation Function (ACF) of Train set")


pacf(y_train, main = "Partial Autocorrelation Function (PACF) of Train set")

```

```{r}
myLB= function(x.fit){
  res=NULL
  npar= dim(x.fit$var.coef)[1]
for (i in (npar+1):40){
  q=Box.test(x.fit$residuals,lag=i,type="Ljung-Box",fitdf=npar)
  res[i]=q$p.value}
  return(res)}
```

#for ARIMA(0,1,2)(0,0,2)[12] 
```{r}
par(mfrow=c(2,2), mar=c(3,3,4,2))
Acf(model_arima_sazonal$residuals, type='correlation', na.action=na.omit, ylab="", main=expression(paste("ACF for Residuals")))
Acf(model_arima_sazonal$residuals, type='partial',  na.action=na.omit, ylab="", main=expression(paste("PACF Residuals")))
plot(myLB(model_arima_sazonal),ylim=c(0,1))
abline(h=0.05,col="blue",lty=2)
```

#for ARIMA(2,1,2)(1,1,2)[12] --> the choosen model that considers a AR component seasonal and not seasonal
```{r}
par(mfrow=c(2,2), mar=c(3,3,4,2))
Acf(model2$residuals, type='correlation', na.action=na.omit, ylab="", main=expression(paste("ACF for Residuals")))
Acf(model2$residuals, type='partial',  na.action=na.omit, ylab="", main=expression(paste("PACF Residuals")))
plot(myLB(model2),ylim=c(0,1))
abline(h=0.05,col="blue",lty=2)
```

```{r}
# Fazer previsões para o conjunto de teste
previsoes_teste <- forecast(model_arima_sazonal, h = length(y_test))
previsoes_teste2 <- forecast(model2, h = length(y_test))

# Plotar as previsões e os dados reais
plot(previsoes_teste, main = "Previsões para o Conjunto de Teste", ylab = "Valores", xlab = "Períodos")

plot(previsoes_teste2, main = "Previsões para o Conjunto de Teste", ylab = "Valores", xlab = "Períodos")
```


```{r}
#resultados <- accuracy(previsoes_teste, y_test)
resultados2 <- accuracy(previsoes_teste2, y_test)

# Visualize as métricas de erro
#print(resultados)
print(resultados2)
```

```{r}
par(mfrow=c(1,1), cex=0.7, mar=c(4,6,1,1))
plot(previsoes_teste, xlim=c(2006,2014))
lines(my_log_data)

par(mfrow=c(1,1), cex=0.7, mar=c(4,6,1,1))
plot(previsoes_teste2, xlim=c(2006,2014))
lines(my_log_data)
```

#Now we build the confidence intervals
```{r}
intervalos_confianca <- c(previsoes_teste$lower, previsoes_teste$upper)

# Visualizar os intervalos de confiança
print(intervalos_confianca)
plot(x = seq_along(previsoes_teste$mean), y = previsoes_teste$mean, type = "l",
     main = "Intervalos de Confiança para Previsões no Set de Teste",
     ylab = "Valores", xlab = "Períodos", ylim = range(c(previsoes_teste$lower, previsoes_teste$upper)))

# Adicionar intervalos de confiança como linhas pontilhadas
lines(x = seq_along(previsoes_teste$mean), y = previsoes_teste$lower[,1], col = "red", lty = 2)
lines(x = seq_along(previsoes_teste$mean), y = previsoes_teste$upper[,1], col = "red", lty = 2)

# Adicionar legenda
legend("topright", legend = c("Previsões", "Intervalo de Confiança"), col = c("black", "red"), lty = c(1, 2))

```

```{r}

intervalos_confianca2 <- c(previsoes_teste2$lower, previsoes_teste2$upper)

# Visualizar os intervalos de confiança
print(intervalos_confianca2)
plot(x = seq_along(previsoes_teste2$mean), y = previsoes_teste2$mean, type = "l",
     main = "Intervalos de Confiança para Previsões no Set de Teste",
     ylab = "Valores", xlab = "Períodos", ylim = range(c(previsoes_teste$lower, previsoes_teste$upper)))

# Adicionar intervalos de confiança como linhas pontilhadas
lines(x = seq_along(previsoes_teste2$mean), y = previsoes_teste2$lower[,1], col = "red", lty = 2)
lines(x = seq_along(previsoes_teste2$mean), y = previsoes_teste2$upper[,1], col = "red", lty = 2)

# Adicionar legenda
legend("topright", legend = c("Previsões", "Intervalo de Confiança"), col = c("black", "red"), lty = c(1, 2))
```

```{r}
# Selecionar os limites inferior e superior dos intervalos de confiança
limites_inferiores <- previsoes_teste$lower
limites_superiores <- previsoes_teste$upper

# Calcular a amplitude de cada intervalo de confiança
amplitude_intervalos <- limites_superiores - limites_inferiores

# Calcular a média e desvio padrão das amplitudes
media_amplitude <- mean(amplitude_intervalos)
desvio_padrao_amplitude <- sd(amplitude_intervalos)

# Imprima os resultados
print("Média da amplitude dos intervalos de confiança:")
print(media_amplitude)
print("Desvio padrão da amplitude dos intervalos de confiança:")
print(desvio_padrao_amplitude)
```


```{r}
# Selecionar os limites inferior e superior dos intervalos de confiança
limites_inferiores2 <- previsoes_teste2$lower
limites_superiores2 <- previsoes_teste2$upper

# Calcular a amplitude de cada intervalo de confiança
amplitude_intervalos2 <- limites_superiores2 - limites_inferiores2

# Calcular a média e desvio padrão das amplitudes
media_amplitude2 <- mean(amplitude_intervalos2)
desvio_padrao_amplitude2 <- sd(amplitude_intervalos2)

# Imprimir os resultados
print("Média da amplitude dos intervalos de confiança:")
print(media_amplitude2)
print("Desvio padrão da amplitude dos intervalos de confiança:")
print(desvio_padrao_amplitude2)
```